Quiz 4
========================================================

Q1
--

```{r}
movies <- read.delim("./data/movies.txt",sep="\t",header=TRUE)
names(movies)
head(movies)
plot(movies$score,movies$box.office,pch=19,col="blue")
lm1 <- lm(movies$score ~ movies$box.office)
lines(movies$box.office,lm1$fitted,col="red",lwd=3)
summary(lm1)
```

Intercept: 50.07559
Slope: 0.09676

The regression coefficient is 0.09676. The interpretation is that an increase of one million dollars in box office gross is associated with a 0.09676 increase in Rotten Tomatoes Score.

Q2
--

```{r}
movies <- read.delim("./data/movies.txt",sep="\t",header=TRUE)
plot(movies$score,movies$box.office,pch=19,col="blue")
lm1 <- lm(movies$score ~ movies$box.office)
lines(movies$box.office,lm1$fitted,col="red",lwd=3)
summary(lm1)
confint(lm1,level=0.9)
```

The 90% confidence interval for the intercept is (47.52, 52.63). If we repeated this study 100 times, we would expect our calculated interval to cover the true value on average 90% of the time.

Q3
--

```{r}
movies <- read.delim("./data/movies.txt",sep="\t",header=TRUE)
par(mfrow=c(1,1))
lmBoth <- lm(movies$score ~ movies$box.office + movies$running.time)
summary(lmBoth)
```
running.time coeff 0.12752

The coefficient is 0.12752. That means that if two movies have the same box office gross, an increase of one minute in running time is associated with an average increase of 0.12752 in score.

Q4
--

Q5
--

Q6
--

Q7
--

Q8
--

Q9
--

Q10
---








