Quiz 4
========================================================

Q1
--

```{r}
movies <- read.delim("./data/movies.txt",sep="\t",header=TRUE)
names(movies)
head(movies)
plot(movies$score,movies$box.office,pch=19,col="blue")
lm1 <- lm(movies$score ~ movies$box.office)
lines(movies$box.office,lm1$fitted,col="red",lwd=3)
summary(lm1)
```

Intercept: 50.07559
Slope: 0.09676

The regression coefficient is 0.09676. The interpretation is that an increase of one million dollars in box office gross is associated with a 0.09676 increase in Rotten Tomatoes Score.

Q2
--

```{r}
movies <- read.delim("./data/movies.txt",sep="\t",header=TRUE)
plot(movies$score,movies$box.office,pch=19,col="blue")
lm1 <- lm(movies$score ~ movies$box.office)
lines(movies$box.office,lm1$fitted,col="red",lwd=3)
summary(lm1)
confint(lm1,level=0.9)
```

The 90% confidence interval for the intercept is (47.52, 52.63). If we repeated this study 100 times, we would expect our calculated interval to cover the true value on average 90% of the time.

Q3
--

```{r}
movies <- read.delim("./data/movies.txt",sep="\t",header=TRUE)
par(mfrow=c(1,1))
lmBoth <- lm(movies$score ~ movies$box.office + movies$running.time)
summary(lmBoth)
```

running.time coeff 0.12752

The coefficient is 0.12752. That means that if two movies have the same box office gross, an increase of one minute in running time is associated with an average increase of 0.12752 in score.

Q4
--

```{r}
movies <- read.delim("./data/movies.txt",sep="\t",header=TRUE)
```

running time is correlated with score

```{r}
anova(lm(movies$score ~ movies$running.time))
```

running time is correlated with box office gross

```{r}
anova(lm(movies$box.office ~ movies$running.time))
```

```{r}
par(mfrow=c(1,2))
plot(movies$running.time,movies$box.office,pch=19,col="blue")
plot(movies$running.time,movies$score,pch=19,col="blue")
```

Yes running time is a confounder. It is correlated both with the Rotten Tomatoes score and the box office gross.

Q5
--

Outliers are present - data points that do not appear to follow the pattern of the other data points

In this case > 175 running time

```{r}
par(mfrow=c(1,2))
movies <- read.delim("./data/movies.txt",sep="\t",header=TRUE)
lmBoth <- lm(movies$score ~ movies$box.office + movies$running.time)
plot(movies$score,movies$running.time,pch=19,col="blue")
# remove the outliers
movies2 = movies[movies$running.time < 175,]
lmBoth2 <- lm(movies2$score ~ movies2$box.office + movies2$running.time)
plot(movies2$score,movies2$running.time,pch=19,col="blue")
summary(lmBoth)
summary(lmBoth2)
```

Yes there are two outliers. After removing them and refitting the regression line, the running time coefficient has a larger magnitude and is more statistically significant.

Q6
--

```{r}
par(mfrow=c(1,2))
movies <- read.delim("./data/movies.txt",sep="\t",header=TRUE)
lmBoth <- lm(movies$score ~ movies$box.office + movies$running.time)
summary(lmBoth)
```

0.0187

The P-value is 0.0187. It is the probability there is no relationship between Rotten Tomatoes score and running time for a fixed box office gross.

Q7
--

```{r}
par(mfrow=c(1,1))
movies <- read.delim("./data/movies.txt",sep="\t",header=TRUE)
lmBoth <- lm(movies$score ~ movies$rating + movies$running.time + movies$running.time*movies$rating)
summary(lmBoth)
```

-0.6901

Q8
--

-0.6901

Q9
--

```{r}
data(warpbreaks)
head(warpbreaks)
lm1 <- lm(warpbreaks$breaks ~ as.factor(warpbreaks$tension))
summary(lm1)
confint(lm1, level=0.95)
```

-17.95042 -2.049581

Q10
---

There is a statistically significant association. We may be skeptical this association could be used to reduce autism rates, since there are many possible explanations for the association that do not involve a direct relationship between organic foods and autism.







