Exploration
========================================================

Obtain data from https://spark-public.s3.amazonaws.com/dataanalysis/loansData.csv

```{r}
url <- "https://spark-public.s3.amazonaws.com/dataanalysis/loansData.csv"
loansFile = "loansData.csv"
# download.file(url, loansFile, method="curl")
```

Load csv into data.frame.

```{r}
loans <- read.csv(loansFile)
names(loans)
head(loans)
```

2500 obs. of 14 variables

FICO.Range is expressed as a range, e.g. 695-699
Employment.Length values: < 1 year, 1 year, ...., 10 years, 10+ years

Clean data

```{r}
lowercols<-function(df){colnames(df)<-tolower(colnames(df))}
colnames(loans)<-lowercols(loans)
fico.split <- strsplit(as.character(loans$fico.range),"\\-")
loans <- transform(loans, fico.low=as.numeric(sapply(fico.split,"[[",1)), fico.high=as.numeric(sapply(fico.split,"[[",2)))
# convert Interest.Rate to numeric
loans$interest.rate <- as.numeric(sub("%", "", loans$interest.rate))
# convert Debt.To.Income.Ration to numeric
loans$debt.to.income.ratio <- as.numeric(sub("%", "", loans$debt.to.income.ratio))
head(loans)
```

fico is left skewed

```{r}
par(mfrow = c(1, 2))
hist(loans$fico.low,col="blue",breaks=100)
hist(log(loans$fico.low+1),col="blue",breaks=100)
```

Strong correlation between interest rate and fico
Low fico -> high interest rate
Number of outliers with higher rates

```{r}
par(mfrow = c(1, 2))
plot(loans$interest.rate ~ loans$fico.range)
lmLow <- lm(loans$interest.rate ~ loans$fico.low)
plot(loans$interest.rate ~ loans$fico.low)
lines(loans$fico.low, lmLow$fitted, col = "red", lwd = 3)
summary(lmLow)
```

High F value low P -> strong correlation

```{r}
anova(lmLow)
```

colnames(loans)
par(mfrow = c(1, 1))

explorereg <- function(df, loanscolumn) {
  plot(df$interest.rate ~ loanscolumn)
  lm1 <- lm(df$interest.rate ~ df$fico.low + loanscolumn)
  summary(lm1)
}

explorereg(loans$debt.to.income.ratio)
explorereg(loans, loans$state)

Some income outliers
Strong correlations when fico is fixed for:
  monthly.income
  employment.length
  inquiries.in.the.last.6.months

summary(loans$monthly.income)
hist(loans$monthly.income, col="blue", breaks=100)
hist(loans$monthly.income[loans$monthly.income < 3e+04], col="blue", breaks=100)
loans.clean <- loans[loans$monthly.income < 3e+04,]
explorereg(loans.clean, loans.clean$monthly.income)  # ***
explorereg(loans, loans$monthly.income)  # ***
explorereg(loans,loans$fico.high)
explorereg(loans,loans$home.ownership) # RENT *
explorereg(loans,loans$open.credit.lines)
explorereg(loans,loans$revolving.credit.balance) # .
explorereg(loans, loans$employment.length) # 10+ years ***
explorereg(loans, loans$inquiries.in.the.last.6.months) # ***
colnames(loans)


